Racets: Policy{-}Agnostic Programming in Racket

### 2018-07-23 paper/paper.tex
recent work on 1

\emph{Policy-agnostic programming} is a linguistic paradigm that
decouples the implementation of privacy policies from the code that
operates on sensitive data. This frees developers to write programs as
they would normally, without inserting logic to manage the policy
directly into application code. Instead, data is labeled with its
policy as it enters the system, and propagates through the program as
computation progresses. Policy-agnostic programming is often
implemented using faceted execution, a dynamic information-flow
monitor encoding sensitive values as \emph{facets}: decision trees
specifying different views of that data. For example, the facet
\facet{Alice}{\code{#t}}{\code{#f}} represents a value that should
appear to Alice as \code{#t} and to everyone but Alice as \blue{\code{#f}}.

Faceted execution propagates facets by extending core linguistic
primitives (such as function application). For example, consider the
application \code{(not x)} where \code{x} is
\blue{\facet{Alice}{\code{#t}}{\code{#f}}}. The programmer intends the
call to \code{not} to be operating on a \code{boolean?}, but because
\code{x} is a facet this call produces \code{#f} instead. The correct
way to interpret function application on facets is to distribute the
application through each branch, producing \zw{producing ...?} %% \code{m
  %% \mc{ \facet{Alice}{\code{#f}}{\code{not #t}}}}
. Other core forms
(such as \code{if}, \code{cond}, etc...) require similar changes.

The core linguistic changes required by faceted execution have led to
relatively inelegant and heavyweight implementations thusfar. For
example, Flanagan et. al (who first presented faceted execution)
extended the JavaScript runtime to account for faceted
values. Similarly, the Jeeves programming language does (what does it
do?)\zw{Need to complete this sentence}. By contrast, Racket's 
highly-expressive macro system allows even
core forms (such as \code{#\%app}) to be redefined with relative ease.

@@

\begin{lstlisting}[language=Racket,escapechar=|,name=example]
(define (mark-hit board x y)
  (if (null? board)
      (cons board #f)
      (let* ([fst (car board)]
             [rst (cdr board)])
        (if (and (= (car fst) x)
                 (= (cdr fst) y))
            (cons rst #t)
            (let ([rst+b (mark-hit rst x y)])
              (cons (cons fst
                          (car rst+b))
                    (cdr rst+b)))))))
\end{lstlisting}

@@

\begin{figure}
\begin{displaymath}
  \begin{array}{lcrcl}
    \blue{c} & \blue{\in} & \blue{const} & \blue{::=} & \blue{() \mid \code{\#t} \mid \code{\#f} \mid \ldots} \\
    \blue{x} & \blue{\in} & \blue{var} & \blue{::=} & \blue{\ldots}  \\
    \blue{e} & \blue{\in} & \blue{exp} & \blue{::=} & \blue{c} \mid \blue{x} \\
    & & & \blue{\mid} & \blue{(\lambda ~ (x) ~ e) \mid (e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{box} ~ e) \mid (\code{unbox} ~ e) \mid (\code{set!} ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{let-label} ~ \ell ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{facet} ~ e ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{obs} ~ e ~ e ~ e)} \\
  \end{array}
\end{displaymath}
\label{fig:syntax}
\caption{Syntax of \lang}
\end{figure}

@@

\begin{lstlisting}[language=Racket,escapechar=|,name=rebindingexample]
(define alice-label (let-label l (lambda (x) (= x alice)) l))
(define x (fac alice-label 1 0))
(let ([alice-label (let-label l (lambda (x) #t) l)]) |\label{lineno:alicelab}|
  (obs alice-label 1 x)) ; Should return |\facet{\texttt{alice-label}}{1}{0}|
\end{lstlisting}

@@

At first glance, it may not be obvious that we even \emph{need} a macro
for facet creation. But according to our semantics, the following
snippet should produce \code{#t} if \bmth{+\ell \in pc}:

\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=other,numbers=none]
(fac l #t (error ''this shouldn't get evaluated if <@\bmth{+\ell \in pc}@>''))
\end{lstlisting}

>>>>

\emph{Policy-agnostic programming} is a linguistic paradigm that
decouples the implementation of privacy policies from the code that
operates on sensitive data. This frees developers to write programs as
they would normally, without inserting logic to manage the policy
directly into application code. Instead, data is labeled with its
policy as it enters the system, and policies propagate through the
program alongside data as computation progresses. Policy-agnostic
programming is often implemented using \emph{faceted execution}, a
dynamic information-flow monitor encoding sensitive values as
\emph{facets}: decision trees specifying different views of that
data. For example, the facet \facet{Alice}{\code{true}}{\code{false}}
represents a value that should appear to Alice as \blue{\code{true}} and to
everyone but Alice as \blue{\code{false}}.

Faceted execution propagates facets by extending core linguistic
primitives (such as function application) to handle faceted
values. For example, consider the application \blue{\code{(x true)}} where
\code{x} is \blue{\facet{Alice}{\lambda x.~\code{true}}{\code{not}}}.
Racket's standard function application will fail here because Racket's
\blue{\code{ppapp}} expects a procedure rather than a facet.  The correct way
to interpret function application on facets is to distribute the
application through each branch, producing
\facet{Alice}{\code{true}}{\code{false}}. Many other core forms (such
as \code{if}, \code{set!}, etc...) require similar changes.

Policy-agnostic programming promotes the idea that programmers should
be able to write programs ``normally'', without concerning themselves
with how privacy policies are enforced. And yet, we believe that the
relatively foundational linguistic changes required to enable faceted
execution have hindered implementations thusfar. For example, the
first implementation of faceted execution (by Austin et
al.~\cite{Austin:12}) extended a JavaScript interpreter to account for
faceted values. Other implementations use monads~\cite{Schmitz:2016}
or rely upon third-party macro systems~\cite{Yang:2016}. So far, we
know of no implementation of faceted execution that works seamlessly
alongside the language. By contrast, Racket has a powerful macro
system, allowing essentially any linguistic form (including function
application) to be extended with arbitrary semantics.

@@

\begin{lstlisting}[language=Racket,escapechar=|,name=example]
(define (mark-hit board x y)
  (if (null? board)
      (cons board false)
      (let* ([fst (car board)]
             [rst (cdr board)])
        (if (and (= (car fst) x)
                 (= (cdr fst) y))
            (cons rst true)
            (let ([rst+b (mark-hit rst x y)])
              (cons (cons fst
                          (car rst+b))
                    (cdr rst+b)))))))
\end{lstlisting}

@@

\begin{figure}
\begin{displaymath}
  \begin{array}{lcrcl}
    \blue{c} & \blue{\in} & \blue{const} & \blue{::=} & \blue{() \mid \code{true} \mid \code{false} \mid \ldots} \\
    \blue{x} & \blue{\in} & \blue{var} & \blue{::=} & \blue{\ldots}  \\
    \blue{e} & \blue{\in} & \blue{exp} & \blue{::=} & \blue{c} \mid \blue{x} \\
    & & & \blue{\mid} & \blue{(\lambda ~ (x) ~ e) \mid (e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{box} ~ e) \mid (\code{unbox} ~ e) \mid (\code{set!} ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{let-label} ~ \ell ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{facet} ~ e ~ e ~ e)} \\
    & & & \blue{\mid} & \blue{(\code{obs} ~ e ~ e ~ e)} \\
  \end{array}
\end{displaymath}
\label{fig:syntax}
\caption{Syntax of \lang}
\end{figure}

@@

\begin{lstlisting}[language=Racket,escapechar=|,name=rebindingexample]
(define alice-label (let-label l (lambda (x) (= x alice)) l))
(define x (fac alice-label 1 0))
(let ([alice-label (let-label l (lambda (x) true) l)]) |\label{lineno:alicelab}|
  (obs alice-label 1 x)) ; Should return |\facet{\texttt{alice-label}}{1}{0}|
\end{lstlisting}

@@

At first glance, it may not be obvious that we even \emph{need} a macro
for facet creation. But according to our semantics, the following
snippet should produce \code{true} if \bmth{+\ell \in pc}:

\begin{lstlisting}[language=Racket,escapeinside={<@}{@>},name=other,numbers=none]
(fac l true (error ''this shouldn't get evaluated if <@\bmth{+\ell \in pc}@>''))
\end{lstlisting}
